---
title: "GSE188236: miRNA target enrichment analysis - the effect of background list"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: TBA

## Introduction

Most popular tools that do enrichment analysis of microRNA target genes don't accept a background list.
Let's see what happens when we use a background list for this type of analysis.
GSE188236 was used because it has both miR and mRNA datasets from the same samples, which means we can
generate a background for the microRNA target gene pathway enrichment analysis.
This dataset examines the effect of TGFB1 expression and co-treatment with SB431542 is a selective and
potent inhibitor of the TGF-beta/Activin/Nodal pathway.
For the sake of this analysis we will only be looking at the comparison of control and TGFB1 cells,
with 5 replicates.

```{r,lib}

library("DESeq2")
library("plyr")
library("dplyr")

```

## Method overview

Load count data from GEO, conduct DESeq2 and make short list of up and downregulated genes.
Select target mRNAs using a database.
Discard any target mRNAs that aren't expressed at sufficient level to be differential.
Conduct enrichment analysis on these target mRNAs.
Repeat the above but without the background list.

## Load data

```{r,load}

if ( file.exists("GSE188236_miRcounts.rds") )

df_mir <- readRDS("GSE188236_miRcounts.rds")

df_mrna <- readRDS("GSE188236_mRNAcounts.rds")

} else {
  download.file(""https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE188236&format=file",
  destfile="GSE188236.tar")
  untar("GSE188236.tar")

  files_mrna <- c("GSM5673193_CTRL_1.txt.gz","GSM5673194_CTRL_2.txt.gz","GSM5673195_CTRL_3.txt.gz",
    "GSM5673196_CTRL_4.txt.gz","GSM5673197_CTRL_5.txt.gz","GSM5673198_TGFB1_1.txt.gz",
    "GSM5673199_TGFB1_2.txt.gz","GSM5673200_TGFB1_3.txt.gz","GSM5673201_TGFB1_4.txt.gz",
    "GSM5673202_TGFB1_5.txt.gz")

  dat_mrna <- lapply(files_mrna,function(f) {
    read.table(f, header=TRUE)
  })

  df_mrna <- join_all(dat_mrna)

  rownames(df_mrna) <- df_mrna$Ensembl_gene_id ; df_mrna$Ensembl_gene_id=NULL

  saveRDS(object=df_mrna,file="GSE188236_mRNAcounts.rds")

  files_mir <- c("GSM5673208_CTRL_1.txt.gz", "GSM5673209_CTRL_2.txt.gz", "GSM5673210_CTRL_3.txt.gz",
    "GSM5673211_CTRL_4.txt.gz", "GSM5673212_CTRL_5.txt.gz", "GSM5673213_TGFB1_1.txt.gz",
    "GSM5673214_TGFB1_2.txt.gz", "GSM5673215_TGFB1_3.txt.gz", "GSM5673216_TGFB1_4.txt.gz",
    "GSM5673217_TGFB1_5.txt.gz")

  dat_mir <- lapply(files_mir,function(f) {
    read.table(f, header=TRUE)
  })

  df_mir <- join_all(dat_mir)

  rownames(df_mir) <- df_mir$Ensembl_ID ; df_mir$Ensembl_ID=NULL

  saveRDS(object=df_mir,file="GSE188236_miRcounts.rds")

  dir.create("GSE188236")

  delfiles <- list.files(".",pattern="txt.gz")

  file.copy(delfiles,"GSE188236",overwrite=TRUE)

   unlink(delfiles)
}

```

## Filtering

```{r,filtering}

dim(df_mir)
summary(rowMeans(df_mir))
df_mir_f <- df_mir[which(rowMeans(df_mir)>9),]
dim(df_mir_f)
df_mir_f <- round(df_mir_f)

dim(df_mrna)
summary(rowMeans(df_mrna))
df_mrna_f <- df_mrna[which(rowMeans(df_mrna)>9),]
dim(df_mrna_f)
summary(rowMeans(df_mrna_f))

```

## Differential analysis

```{r,deseq}

ss <- data.frame(colnames(df_mir_f))
rownames(ss) <- ss[,1]
colnames(ss) <- "samplename"
ss$trt <- factor(grepl("TGF",ss$samplename))

dds <- DESeqDataSetFromMatrix(countData = df_mir_f , colData = ss, design = ~ trt )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE,nsub=200)
zz <- cbind(as.data.frame(z),assay(vsd))
dge_mir <- as.data.frame(zz[order(zz$pvalue),])
dge_mir[1:20,1:6] %>%
  kbl(caption = "Top gene expression differences between ctrl and TGFB1 treated cells") %>%
  kable_paper("hover", full_width = F)




```


## Conclusions


