---
title: "GSE188236: miRNA target enrichment analysis - the effect of background list"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: TBA

## Introduction

Most popular tools that do enrichment analysis of microRNA target genes don't accept a background list.
Let's see what happens when we use a background list for this type of analysis.
GSE188236 was used because it has both miR and mRNA datasets from the same samples, which means we can
generate a background for the microRNA target gene pathway enrichment analysis.
This dataset examines the effect of TGFB1 expression and co-treatment with SB431542 is a selective and
potent inhibitor of the TGF-beta/Activin/Nodal pathway.
For the sake of this analysis we will only be looking at the comparison of control and TGFB1 cells,
with 5 replicates.

```{r,lib}

library("DESeq2")
library("plyr")
library("dplyr")
library("kableExtra")
library("readxl")
library("fgsea")
library("eulerr")
```

## Method overview

Load count data from GEO, conduct DESeq2 and make short list of up and downregulated genes.
Select target mRNAs using a database.
Discard any target mRNAs that aren't expressed at sufficient level to be differential.
Conduct enrichment analysis on these target mRNAs.
Repeat the above but without the background list.

## Load data

```{r,load}

if ( file.exists("GSE188236_miRcounts.rds") ) {

  df_mir <- readRDS("GSE188236_miRcounts.rds")

  df_mrna <- readRDS("GSE188236_mRNAcounts.rds")

} else {
  download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE188236&format=file",
  destfile="GSE188236.tar")
  untar("GSE188236.tar")

  files_mrna <- c("GSM5673193_CTRL_1.txt.gz","GSM5673194_CTRL_2.txt.gz","GSM5673195_CTRL_3.txt.gz",
    "GSM5673196_CTRL_4.txt.gz","GSM5673197_CTRL_5.txt.gz","GSM5673198_TGFB1_1.txt.gz",
    "GSM5673199_TGFB1_2.txt.gz","GSM5673200_TGFB1_3.txt.gz","GSM5673201_TGFB1_4.txt.gz",
    "GSM5673202_TGFB1_5.txt.gz")

  dat_mrna <- lapply(files_mrna,function(f) {
    read.table(f, header=TRUE)
  })

  df_mrna <- join_all(dat_mrna)

  rownames(df_mrna) <- df_mrna$Ensembl_gene_id ; df_mrna$Ensembl_gene_id=NULL

  saveRDS(object=df_mrna,file="GSE188236_mRNAcounts.rds")

  files_mir <- c("GSM5673208_CTRL_1.txt.gz", "GSM5673209_CTRL_2.txt.gz", "GSM5673210_CTRL_3.txt.gz",
    "GSM5673211_CTRL_4.txt.gz", "GSM5673212_CTRL_5.txt.gz", "GSM5673213_TGFB1_1.txt.gz",
    "GSM5673214_TGFB1_2.txt.gz", "GSM5673215_TGFB1_3.txt.gz", "GSM5673216_TGFB1_4.txt.gz",
    "GSM5673217_TGFB1_5.txt.gz")

  dat_mir <- lapply(files_mir,function(f) {
    read.table(f, header=TRUE)
  })

  df_mir <- join_all(dat_mir)

  rownames(df_mir) <- df_mir$Ensembl_ID ; df_mir$Ensembl_ID=NULL

  saveRDS(object=df_mir,file="GSE188236_miRcounts.rds")

  dir.create("GSE188236")

  delfiles <- list.files(".",pattern="txt.gz")

  file.copy(delfiles,"GSE188236",overwrite=TRUE)

   unlink(delfiles)
}

```

## Filtering

```{r,filtering}

dim(df_mir)
summary(rowMeans(df_mir))
df_mir_f <- df_mir[which(rowMeans(df_mir)>9),]
dim(df_mir_f)
df_mir_f <- round(df_mir_f)

dim(df_mrna)
summary(rowMeans(df_mrna))
df_mrna_f <- df_mrna[which(rowMeans(df_mrna)>9),]
dim(df_mrna_f)
summary(rowMeans(df_mrna_f))

```

## Load ensembl 105 data

Not sure whether this is the correct one.
Just a guess.
There are 60666 rows of data and 60531 in the v105 annotation which is quite close.
Data was downloaded from Ensembl biomart archive 27th Nov 2023.

```{r,anno}

gt <- read.table("mart_export.txt",sep="\t",header=TRUE)

```

## Differential analysis miRNA

```{r,deseq1}

ss <- data.frame(colnames(df_mir_f))
rownames(ss) <- ss[,1]
colnames(ss) <- "samplename"
ss$trt <- factor(grepl("TGF",ss$samplename))

dds <- DESeqDataSetFromMatrix(countData = df_mir_f , colData = ss, design = ~ trt )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE,nsub=200)
zz <- cbind(as.data.frame(z),assay(vsd))
dge_mir <- as.data.frame(zz[order(zz$pvalue),])

dge_mir[1:20,1:6] %>%
  kbl(caption = "Top miRNA expression differences between ctrl and TGFB1 treated cells") %>%
  kable_paper("hover", full_width = F)

mir_up <- rownames(subset(dge_mir,padj<0.05 & log2FoldChange >0 ))
mir_up
mir_up <- gt[which(gt$Gene.stable.ID.version %in% mir_up),]
mir_up$Gene.name <- tolower(mir_up$Gene.name)
mir_up

mir_dn <- rownames(subset(dge_mir,padj<0.05 & log2FoldChange <0 ))
mir_dn
mir_dn <- gt[which(gt$Gene.stable.ID.version %in% mir_dn),]
mir_dn
mir_dn$Gene.name <- tolower(mir_dn$Gene.name)
mir_dn

```

## Differential analysis mRNA

```{r,deseq2}

ss <- data.frame(colnames(df_mrna_f))
rownames(ss) <- ss[,1]
colnames(ss) <- "samplename"
ss$trt <- factor(grepl("TGF",ss$samplename))

dds <- DESeqDataSetFromMatrix(countData = df_mrna_f , colData = ss, design = ~ trt )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE,nsub=200)
zz <- cbind(as.data.frame(z),assay(vsd))
dge_mrna <- as.data.frame(zz[order(zz$pvalue),])

dge_mrna[1:20,1:6] %>%
  kbl(caption = "Top mRNA gene expression differences between ctrl and TGFB1 treated cells") %>%
  kable_paper("hover", full_width = F)

mrna_up <- rownames(subset(dge_mrna,padj<0.05 & log2FoldChange >0 ))
mrna_up
mrna_up <- gt[which(gt$Gene.stable.ID %in% mrna_up),]
mrna_up <- unique(mrna_up$Gene.name)
mrna_up <- mrna_up[mrna_up != ""]
mrna_up

mrna_dn <- rownames(subset(dge_mrna,padj<0.05 & log2FoldChange <0 ))
mrna_dn
mrna_dn <- gt[which(gt$Gene.stable.ID %in% mrna_dn),]
mrna_dn <- unique(mrna_dn$Gene.name)
mrna_dn <- mrna_dn[mrna_dn != ""]
mrna_dn

```

## Fetch target gene database

```{r,miRtarbase}

#options(timeout=1000)
#download.file("https://mirtarbase.cuhk.edu.cn/~miRTarBase/miRTarBase_2022/cache/download/9.0/hsa_MTI.xlsx",
#destfile="hsa_MTI.xlsx",)

mirtarbase <- read_excel("hsa_MTI.xlsx")

# mir up
mir_up_gn <- mir_up$Gene.name
mir_up_gn2 <- gsub("$","-",gsub("mir","hsa-miR-",mir_up_gn))
mir_up_gn2 <- gsub("a2","a",mir_up_gn2)

up_targets <- lapply(mir_up_gn2,function(x) {
  a <- unique(unlist(as.vector(mirtarbase[grep(x,mirtarbase$miRNA),"Target Gene"])))
  y <- gsub("-$","",x)
  b <- unique(unlist(as.vector(mirtarbase[which(mirtarbase$miRNA %in% y),"Target Gene"])))
  d <- unique(c(a,b))
  return(d)
})

lapply(up_targets,length)
up_targets <- unique(unlist(up_targets))
up_targets
length(up_targets)

# mir dn
mir_dn_gn <- mir_dn$Gene.name
mir_dn_gn2 <- gsub("$","-",gsub("mir","hsa-miR-",mir_dn_gn))

dn_targets <- lapply(mir_dn_gn2,function(x) {
  a <- unique(unlist(as.vector(mirtarbase[grep(x,mirtarbase$miRNA),"Target Gene"])))
  y <- gsub("-$","",x)
  b <- unique(unlist(as.vector(mirtarbase[which(mirtarbase$miRNA %in% y),"Target Gene"])))
  d <- unique(c(a,b))
  return(d)
})

lapply(dn_targets,length)
dn_targets <- unique(unlist(dn_targets))
dn_targets
length(dn_targets)

```

## miR target Enrichment with FORA

```{r,orabg}

reactome <- gmtPathways("ReactomePathways_2023-11-28.gmt")

kegg <- gmtPathways("c2.cp.kegg_medicus.v2023.2.Hs.symbols.gmt")

summary(unlist(lapply(reactome,length)))

summary(unlist(lapply(kegg,length)))

# make a gene list - some of the ensembl IDs dont have a symbol (~2000 of ~17000)
gl<-gt[gt$Gene.stable.ID %in% rownames(df_mrna_f),"Gene.name"]

# make a non-redundant list of gene symbols
bg <- unique(gl[which(gl!="")])

length(gl)
length(bg)

```

First try with whole genome as background.

```{r,fora_wgbg}

wgbg <- unique(gt$Gene.name)
wgbg <- wgbg[which(wgbg!="")]

setdiff(up_targets,wgbg)
length(setdiff(up_targets,wgbg))
length(setdiff(up_targets,wgbg))/length(up_targets)
up_wg_res <- fora(pathways=reactome, genes=up_targets,  universe=wgbg, minSize = 5)
nbg = length(wgbg)
nsel = length(intersect(up_targets,wgbg))
up_wg_res$foldenrichment <- (up_wg_res$overlap / nsel ) / ( up_wg_res$size / nbg )
head(up_wg_res)
nrow(up_wg_res)
nrow(subset(up_wg_res,padj<0.05 & foldenrichment >2))
up_wg_res_sets <- subset(up_wg_res,padj<0.05 & foldenrichment >2)$pathway

setdiff(dn_targets,wgbg)
length(setdiff(dn_targets,wgbg))
length(setdiff(dn_targets,wgbg))/length(dn_targets)
dn_wg_res <- fora(pathways=reactome, genes=dn_targets,  universe=wgbg, minSize = 5)
nbg = length(wgbg)
nsel = length(intersect(dn_targets,wgbg))
dn_wg_res$foldenrichment <- (dn_wg_res$overlap / nsel ) / ( dn_wg_res$size / nbg )
head(dn_wg_res)
nrow(dn_wg_res)
nrow(subset(dn_wg_res,padj<0.05 & foldenrichment >2))
dn_wg_res_sets <- subset(dn_wg_res,padj<0.05 & foldenrichment >2)$pathway

```

Now try with proper background.

```{r,fora_bg}

setdiff(up_targets,bg)
length(setdiff(up_targets,bg))
length(intersect(up_targets,bg))
length(up_targets)
length(setdiff(up_targets,bg))/length(up_targets)
up_bg_res <- fora(pathways=reactome, genes=up_targets,  universe=bg, minSize = 5)
nbg = length(bg)
nsel = length(intersect(up_targets,bg))
up_bg_res$foldenrichment <- (up_bg_res$overlap / nsel ) / ( up_bg_res$size / nbg )
head(up_bg_res)
nrow(up_bg_res)
nrow(subset(up_bg_res,padj<0.05 & foldenrichment >2))
up_bg_res_sets <- subset(up_bg_res,padj<0.05 & foldenrichment >2)$pathway

setdiff(dn_targets,bg)
length(setdiff(dn_targets,bg))
length(intersect(dn_targets,bg))
length(dn_targets)
length(setdiff(dn_targets,bg))/length(dn_targets)
dn_bg_res <- fora(pathways=reactome, genes=dn_targets,  universe=bg, minSize = 5)
nbg = length(bg)
nsel = length(intersect(dn_targets,bg))
dn_bg_res$foldenrichment <- (dn_bg_res$overlap / nsel ) / ( dn_bg_res$size / nbg )
head(dn_bg_res)
nrow(dn_bg_res)
nrow(subset(dn_bg_res,padj<0.05 & foldenrichment >2))
dn_bg_res_sets <- subset(dn_bg_res,padj<0.05 & foldenrichment >2)$pathway

```


```{r,compare2,fig.height=5,fig.width=5}

v1 <- list("WG up"=up_wg_res_sets,"BG up"=up_bg_res_sets,
  "WG dn"=dn_wg_res_sets,"BG dn"=dn_bg_res_sets )

plot(euler(v1),quantities = TRUE,main="Effect of background list")

```

Now compare miR targets up and down together with observed up and down.

```{r,compare3,fig.height=5,fig.width=5}

v2 <- list("up DGE"=mrna_up, "dn DGE"=mrna_dn,
  "up miR targets"=up_targets, "dn miR targets"=dn_targets)

plot(euler(v2),quantities = TRUE, main="miR targets and DEGs?")

```

Interesting how low the enrichment is.

Now check with 100 genes.

```{r,compare4,fig.height=5,fig.width=5}

mrna_up <- rownames(head(subset(dge_mrna,log2FoldChange >0 ),129))
str(mrna_up)
mrna_up <- unique(gt[which(gt$Gene.stable.ID %in% mrna_up),"Gene.name"])
mrna_up <- mrna_up[mrna_up != ""]
str(mrna_up)

mrna_dn <- rownames(head(subset(dge_mrna,log2FoldChange <0 ),105))
str(mrna_dn)
mrna_dn <- unique(gt[which(gt$Gene.stable.ID %in% mrna_dn),"Gene.name"])
mrna_dn <- mrna_dn[mrna_dn != ""]
str(mrna_dn)

v2 <- list("up DGE"=mrna_up, "dn DGE"=mrna_dn,
  "up miR targets"=up_targets, "dn miR targets"=dn_targets)

plot(euler(v2),quantities = TRUE, main="miR targets and DEGs?")

```

Now check with 1000 genes.

```{r,compare5,fig.height=5,fig.width=5}

mrna_up <- rownames(head(subset(dge_mrna,log2FoldChange >0 ),1288))
str(mrna_up)
mrna_up <- unique(gt[which(gt$Gene.stable.ID %in% mrna_up),"Gene.name"])
mrna_up <- mrna_up[mrna_up != ""]
str(mrna_up)

mrna_dn <- rownames(head(subset(dge_mrna,log2FoldChange <0 ),1080))
str(mrna_dn)
mrna_dn <- unique(gt[which(gt$Gene.stable.ID %in% mrna_dn),"Gene.name"])
mrna_dn <- mrna_dn[mrna_dn != ""]
str(mrna_dn)

v2 <- list("up DGE"=mrna_up, "dn DGE"=mrna_dn,
  "up miR targets"=up_targets, "dn miR targets"=dn_targets)

plot(euler(v2),quantities = TRUE, main="miR targets and DEGs?")

```

Now check with 250 genes.

```{r,compare6}

mrna_up <- rownames(head(subset(dge_mrna,log2FoldChange >0 ),321))
str(mrna_up)
mrna_up <- unique(gt[which(gt$Gene.stable.ID %in% mrna_up),"Gene.name"])
mrna_up <- mrna_up[mrna_up != ""]
str(mrna_up)

mrna_dn <- rownames(head(subset(dge_mrna,log2FoldChange <0 ),266))
str(mrna_dn)
mrna_dn <- unique(gt[which(gt$Gene.stable.ID %in% mrna_dn),"Gene.name"])
mrna_dn <- mrna_dn[mrna_dn != ""]
str(mrna_dn)

```

## Now pathway analysis of long RNA-seq

Use top 250 genes in either direction for enrichment analysis.
There was a paper by Tarca recommending this.

```{r,mrna_fora_bg}

mrna_up_res <- fora(pathways=reactome, genes=mrna_up,  universe=bg, minSize = 5)
nbg = length(bg)
nsel = length(intersect(mrna_up,bg))
mrna_up_res$foldenrichment <- (mrna_up_res$overlap / nsel ) / ( mrna_up_res$size / nbg )
head(mrna_up_res)
nrow(mrna_up_res)
nrow(subset(mrna_up_res,padj<0.05 & foldenrichment >0))
mrna_up_res_sets <- subset(mrna_up_res,padj<0.05 & foldenrichment >1)$pathway

mrna_dn_res <- fora(pathways=reactome, genes=mrna_dn,  universe=bg, minSize = 5)
nbg = length(bg)
nsel = length(intersect(mrna_dn,bg))
mrna_dn_res$foldenrichment <- (mrna_dn_res$overlap / nsel ) / ( mrna_dn_res$size / nbg )
head(mrna_dn_res)
nrow(mrna_dn_res)
nrow(subset(mrna_dn_res,padj<0.05 & foldenrichment >0))
mrna_dn_res_sets <- subset(mrna_dn_res,padj<0.05 & foldenrichment >1)$pathway

```

```{r,euler_pathways,fig.height=5,fig.width=5}

v3 <- list("dn miR target"=dn_bg_res_sets,
  "up miR target"=up_bg_res_sets,
  "up mRNA"=mrna_up_res_sets,
  "dn mRNA"=mrna_dn_res_sets)

plot(euler(v3),quantities = TRUE, main="miR target and DE pathways")

```

## Conclusions


