---
title: "Sampling high impact articles doing DIANA miRpath analysis in pubmed"
author: "Survey enrichment team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/SurveyEnrichment2/pubmed

## Introduction

Enrichment analysis commonly suffers from statistical problems and poor reporting, making the findings
irreproducible.
Here we are investigating the methodology of some high impact articles that have conducted enrichment
analysis.
In order to highlight these problematic articles, we need to collect a corpus of them to screen.
A list of pubmed articles cting DAVID was collected.
The SJR (Scimago Journal Rank) for prominent journals was also collected.
These will be merged and high impact enrichment articles will be curated.

```{r,libs}

library("kableExtra")

```

## Load data

SJR data were downloaded from [here](https://www.scimagojr.com/journalrank.php?min=200&min_type=cd).

Pubmed data were downloaded from the "cited by" pages for the tool's article PMID: 31684860.

The data were downloaded in "Pubmed Format".

The pubmed2tbl script was used to convert this into tabular format.

```{r,load1}

sjr <- read.csv("scimagojr_2022.csv",header=TRUE,sep=";")
colnames(sjr)

```

Need to clean some data.
Removing all data except the ISSN and the PMID.

```{bash,clean1}

# extract PMID and ISSN from PUBMED format data

for TXT in *set.txt ; do

  sed 's/^PMID/!PMID/' $TXT \
  | tr '\r' '\n' \
  | tr -d ' ' \
  | grep -v ^$ \
  | tr '\n' '\t' \
  | tr '!' '\n' > tmp.txt

  for PMID in $(cut -f1 tmp.txt | cut -d '-' -f2 ) ; do

    SEARCH="PMID-$PMID"
    IS=$(grep -w "$SEARCH" tmp.txt \
      | tr '\t' '\n' \
      | grep -m1 ^IS \
      | cut -d '(' -f1 \
      | cut -d '-' -f2- \
      | tr -d '-' )
    echo $SEARCH $IS | sed 's/PMID-//'

  done > $TXT.tsv

  rm tmp.txt

done

```

Now it should be possible to load in to R.

```{r,load2}

pm <- read.table("pubmed-31684860-set.txt.tsv",fill=NA)

dim(pm)

colnames(pm) <- c("PMID","ISSN")

```

## Merge

First need to get the electronic ISSN, then cut down the dataset to
include just the eISSN and the SJR.

```{r,merge1}

rev1 <- lapply(sjr$Issn , function(x) {intToUtf8(rev(utf8ToInt(x)))} )
rev1 <- do.call(rbind,rev1)
rev1 <- sapply(strsplit(rev1," "),"[[",1)
issn <- lapply(rev1 , function(x) {intToUtf8(rev(utf8ToInt(x)))} )
issn <- unlist(issn)
sjr$issn <- issn

sjr2 <- sjr[,c("SJR","issn")]

m <- merge(sjr2,pm,by.x="issn",by.y="ISSN")

m$SJR <- as.numeric(gsub(",",".",m$SJR))

```

```{r,hist1}

hist(m$SJR)
nrow(subset(m,SJR>5))
nrow(subset(m,SJR>10))
nrow(subset(m,SJR>15))

m <- subset(m,SJR>5)

m %>%
  kbl(caption="top high impact articles describing DIANA miRpath analysis") %>%
  kable_paper("hover",full_width=FALSE)

write.table(m,file="corpus_mienturnet.tsv",row.names=FALSE)

```

## Session Information

```{r,session}

sessionInfo()

```

